### 2024-07-28 mixed-models.qmd
20240727 - selective attrition

Based on the bivariate scatterplots between age and fantasy points, we might conclude that players tend to stay stable or even increase in fantasy points with age.
However, we would be wrong.
When we account for the longitudinal data (i.e., multiple observations over time for the same player) using mixed models, we observe that fantasy points tend to decrease with age, with the timing and rate of decline differing for each position.
In other words, the association between age and fantasy points differs at the person level versus the group level.
This is an example of Simpson's paradox.
The discrepancy between the positive or null association between age and fantasy points at the group level, and the negative association at the person level may be due, in part, to the selective attrition of players with age.
The highest performing players will tend to play the longest, whereas the poorest performing players will retire or get dropped from the team at the youngest ages.
Thus, the selective attrition may make it seem that there is no association between age and performance (or even a positive one!), when in fact, players' performance tends to decrease with age after age 26 or so (with the timing differing from position to position).

>>>>

Based on the bivariate scatterplots between age and fantasy points, we might conclude that players tend to stay stable or even increase in fantasy points with age.
However, we would be wrong.
When we account for the longitudinal data (i.e., multiple observations over time for the same player) using mixed models, we observe that fantasy points tend to decrease with age, with the timing and rate of decline differing for each position.
In other words, the association between age and fantasy points differs at the person level versus the group level.
This is an example of Simpson's paradox.

The discrepancy between the positive or null association between age and fantasy points at the group level, and the negative association at the person level may be due, in part, to the selective attrition of players with age.
The highest performing players will tend to play the longest, whereas the poorest performing players will retire or get dropped from the team at the youngest ages.
Thus, the selective attrition may make it seem that there is no association between age and performance (or even a positive one!), when in fact, players' performance tends to decrease with age after age 26 or so (with the timing differing from position to position), until the player eventually retires or is dropped from the team.
Selective attrition is common in longitudinal studies and intervention studies.
For instance, attrition may be more likely for individuals from lower socioeconomic status backgrounds because they may face more challenges in continuing in longitudinal studies such as fewer financial resources, greater life stressors, etc.
In addition, people who experience side effects or lack of improvement may be more likely to drop out of intervention studies.
Examining only those who completed treatment (an example of [selection bias](#sec-causalDiagramSelectionBias)) would make the intervention look more effective than it actually was because the people who stay in the study are those who experience the greatest improvement.
Thus, it is important to use approaches such as mixed models or other approaches that account for the multiple observations from the same person, that use all available information, and that do not exclude people who do not complete all portions of the study.
